### 2013-02-13 inst/doc/nonparametric-bayes.Rmd
writing

## Pseudocode for the determining the transtion matrix from the GP
```r
for(h in h_grid)
  F_h = for(x_j in grid)
          for(i in 1:N) 
            dnorm(x_j, mu[i]-h, V[i])
```

@@

### Discussion of state equation

@@

### Brief discussion on choice of model parameters, nuisance parameters

### Discussion of training data

Both parametric and nonparametric approaches will require some training
data on which to base their model of the process.  We generate the
training data under the model described in Eq 1 for `r Tobs` time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$.  For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the harvest.
(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

We estimate two parametric models from the data using a maximum likelihood
approach.  The first model is structurally identical to the true model
(Eq 1), differing only in that it's parameters are estimated from the
observed data rather than given.  The alternative model is the Ricker
model, which is structurally similar and commonly assumed

>>>>

## Pseudocode for the determining the transtion matrix from the GP

```r
for(h in h_grid)
  F_h = for(x_j in grid)
          for(i in 1:N) 
            dnorm(x_j, mu[i]-h, V[i])
```

@@

## Example in a bistable system

@@

## Sample training data

Both parametric and nonparametric approaches will require training
data on which to base their model of the process.  We generate the
training data under the model described in Eq 1 for `r Tobs` time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$.  For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the harvest.

<!--Should we include any emprical examples? -->
Using data simulated from a specified model rather than empirical data
permits the comparison against the true underlying dynamics, setting 
a bar for the optimal performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

We estimate two parametric models from the data using a maximum likelihood
approach.  The first model is structurally identical to the true model
(Eq 1), differing only in that it's parameters are estimated from the
observed data rather than given.  The alternative model is the Ricker
model, which is structurally similar and commonly used in for such data.

### 2013-02-13 inst/doc/nonparametric-bayes.tex
writing

\subsubsection{Discussion of state equation}

\subsubsection{Brief discussion on choice of model parameters, nuisance
parameters}

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape
parameter $\sigma_g$. We start with an example in which the parameters
are $r =$ \texttt{2}, $K =$ \texttt{10}, $C =$ \texttt{5}, and
$\sigma_g =$ \texttt{0.05}.

Both parametric and nonparametric approaches will require some training
data on which to base their model of the process. We generate the
training data under the model described in Eq 1 for \texttt{35} time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$. For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the
harvest. (Motivation, alternatives, stationarity, examples without a
stable node (limit-cycle models), examples based on observations near a
stable node alone, and why that isn't impossible).

Which estimates a Ricker model with $r =$ \texttt{1.8501}, $K =$
\texttt{9.8091}, and the Allen Allele model with $r =$ \texttt{2.8079},
$K =$ \texttt{11.8235} and $C =$ \texttt{7.2159}.

@@

The posterior Gaussian process is completely defined by the expected
value and covariance matrix at a defined set of training points. For
simplicty we will consider a these points to fall on a discrete, uniform
grid $x$ of \texttt{101} points from \texttt{0} to \texttt{15} (1.5
times the positive equilibrium $K$). Again to keep things simple we will
use this same grid discritization for the parametric approach. Other
options for choosing the grid points, including collocation methods and
functional basis expansion (or even using Guassian processes in place of
the discrete optimization; an entirely different context in which GP can
be used in SDP, see {[}@Deisenroth2009{]}) could also be considered.

The transition matrix $\mathbf{F}$ is thus an \texttt{101} by
\texttt{101} matrix for which the ${i,j}$ entry gives the probability of
transitioning into state $x_i$ given that the system is in state $x_j$
in the previous timestep. To generate the transition matrix based on the
posterior GP, we need only the expected values at each grid point and
the corresponding variances (the diagonal of the covariance matrix), as
shown in Figure 1. Given the mean at each gridpoint as the length
\texttt{101} vector $E$ and variance $V$, the probability of
transitioning from state $x_i$ to state $x_j$ is simply
$\mathcal{N}\left(x_j | \mu = E_i, \sigma = \sqrt{V_i}\right)$, where
$\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{for(h in h_grid)}
  \NormalTok{F_h = for(x_j in grid)}
          \NormalTok{for(i in }\DecValTok{1}\NormalTok{:N) }
            \KeywordTok{dnorm}\NormalTok{(x_j, mu[i]-h, V[i])}
\end{Highlighting}
\end{Shaded}

@@

where $V(x_t)$ is the value of being at state $x$ at time $t$, $h$ is
control (harvest level) chosen. Numerically, the maximization is
accomplished as follows. Consider the set of possible control values to
be the discrete \texttt{101} values corresponding the the grid of stock
sizes. Then for each $h_t$ there is a corresponding transition matrix
$\mathbf{F}_h$ determined as described above but with mean
$\mu = x_j - h_t$. Let $\vec{V_t}$ be the vector whose $i$th element
corresponds to the value of having stock $x_i$ at time $t$. Then let
$\Pi_h$ be the vector whose $i$th element indicates the profit from
harvesting at intensity $h_t$ given a population $x_i$ (e.g.
$\max(x_i, h_t)$ since one cannot harvest more fish then the current
population size). Then the Bellman recursion can be given in matrix form
as

@@

\subsection{Pseudocode for the Bellman iteration}

\begin{Shaded}
\begin{Highlighting}[]
 \NormalTok{V1 <- }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(h_grid), function(h)\{}
      \NormalTok{delta * F[[h]] %*% V +  }\KeywordTok{profit}\NormalTok{(x_grid, h_grid[h]) }
    \NormalTok{\})}
    \CommentTok{# find havest, h that gives the maximum value}
    \NormalTok{out <- }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:gridsize, function(j)\{}
      \NormalTok{value <- }\KeywordTok{max}\NormalTok{(V1[j,], }\DataTypeTok{na.rm =} \NormalTok{T) }\CommentTok{# each col is a diff h, max over these}
      \NormalTok{index <- }\KeywordTok{which.max}\NormalTok{(V1[j,])  }\CommentTok{# store index so we can recover h's }
      \KeywordTok{c}\NormalTok{(value, index) }\CommentTok{# returns both profit value & index of optimal h.  }
    \NormalTok{\})}
    \CommentTok{# Sets V[t+1] = max_h V[t] at each possible state value, x}
    \NormalTok{V <- out[}\DecValTok{1}\NormalTok{,]                        }\CommentTok{# The new value-to-go}
    \NormalTok{D[,OptTime-time}\DecValTok{+1}\NormalTok{] <- out[}\DecValTok{2}\NormalTok{,]       }\CommentTok{# The index positions}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\item
  Replicate stochastic simulations
\item
  Sensitivity analysis (Figure 4).
\end{itemize}

@@

@@

\begin{itemize}
\item
  Multiple species
\item
  Online learning
\item
  Multiple step-ahead predictions
\item
  Explicitly accomidating additional uncertainties
\item
  Improving inference of optimal policy from the GP
\end{itemize}

@@

Allen, Linda J. S., Jesse F. Fagan, Göran Högnäs, and Henrik Fagerholm.
2005. ``Population extinction in discrete-time stochastic population
models with an Allee effect.'' \emph{Journal of Difference Equations and
Applications} 11 (apr): 273--293. doi:10.1080/10236190412331335373.

@@

Cressie, Noel, Catherine a Calder, James S. Clark, Jay M. Ver Hoef, and
Christopher K. Wikle. 2009. ``Accounting for uncertainty in ecological
analysis: the strengths and limitations of hierarchical statistical
modeling.'' \emph{Ecological Applications} 19 (apr): 553--70.

Geritz, Stefan a. H., and Eva Kisdi. 2011. ``Mathematical ecology: why
mechanistic models?.'' \emph{Journal of mathematical biology} (dec).
doi:10.1007/s00285-011-0496-3.

Levins, Richard. 1966. ``The strategy of model building in population
biology.'' \emph{American Scientist} 54: 421--431.

@@

Munch, Stephan B., Melissa L. Snover, George M. Watters, and Marc
Mangel. 2005. ``A unified treatment of top-down and bottom-up control of
reproduction in populations.'' \emph{Ecology Letters} 8 (may): 691--695.
doi:10.1111/j.1461-0248.2005.00766.x.

@@

Reed, William J. 1979. ``Optimal escapement levels in stochastic and
deterministic harvesting models.'' \emph{Journal of Environmental
Economics and Management} 6 (dec): 350--363.
doi:10.1016/0095-0696(79)90014-7.

Scheffer, Marten, Stephen R. Carpenter, J. A. Foley, C. Folke, and B.
Walker. 2001. ``Catastrophic shifts in ecosystems.'' \emph{Nature} 413
(oct): 591--6. doi:10.1038/35098000.

>>>>

@@

The posterior Gaussian process is completely defined by the expected
value and covariance matrix at a defined set of training points. For
simplicty we will consider a these points to fall on a discrete, uniform
grid $x$ of 101 points from 0 to 15 (1.5 times the positive equilibrium
$K$). Again to keep things simple we will use this same grid
discritization for the parametric approach. Other options for choosing
the grid points, including collocation methods and functional basis
expansion (or even using Guassian processes in place of the discrete
optimization; an entirely different context in which GP can be used in
SDP, see {[}@Deisenroth2009{]}) could also be considered.

The transition matrix $\mathbf{F}$ is thus an 101 by 101 matrix for
which the ${i,j}$ entry gives the probability of transitioning into
state $x_i$ given that the system is in state $x_j$ in the previous
timestep. To generate the transition matrix based on the posterior GP,
we need only the expected values at each grid point and the
corresponding variances (the diagonal of the covariance matrix), as
shown in Figure 1. Given the mean at each gridpoint as the length 101
vector $E$ and variance $V$, the probability of transitioning from state
$x_i$ to state $x_j$ is simply
$\mathcal{N}\left(x_j \vert  \mu = E_i, \sigma = \sqrt{V_i}\right)$,
where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{for(h in h_grid)}
  \NormalTok{F_h =}\StringTok{ }\NormalTok{for(x_j in grid)}
          \NormalTok{for(i in }\DecValTok{1}\NormalTok{:N) }
            \KeywordTok{dnorm}\NormalTok{(x_j, mu[i]-h, V[i])}
\end{Highlighting}
\end{Shaded}

@@

where $V(x_t)$ is the value of being at state $x$ at time $t$, $h$ is
control (harvest level) chosen. Numerically, the maximization is
accomplished as follows. Consider the set of possible control values to
be the discrete 101 values corresponding the the grid of stock sizes.
Then for each $h_t$ there is a corresponding transition matrix
$\mathbf{F}_h$ determined as described above but with mean
$\mu = x_j - h_t$. Let $\vec{V_t}$ be the vector whose $i$th element
corresponds to the value of having stock $x_i$ at time $t$. Then let
$\Pi_h$ be the vector whose $i$th element indicates the profit from
harvesting at intensity $h_t$ given a population $x_i$ (e.g.
$\max(x_i, h_t)$ since one cannot harvest more fish then the current
population size). Then the Bellman recursion can be given in matrix form
as

@@

\subsubsection{Pseudocode for the Bellman iteration}

\begin{Shaded}
\begin{Highlighting}[]
 \NormalTok{V1 <-}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(h_grid), function(h)\{}
      \NormalTok{delta *}\StringTok{ }\NormalTok{F[[h]] %*%}\StringTok{ }\NormalTok{V +}\StringTok{  }\KeywordTok{profit}\NormalTok{(x_grid, h_grid[h]) }
    \NormalTok{\})}
    \CommentTok{# find havest, h that gives the maximum value}
    \NormalTok{out <-}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\DecValTok{1}\NormalTok{:gridsize, function(j)\{}
      \NormalTok{value <-}\StringTok{ }\KeywordTok{max}\NormalTok{(V1[j,], }\DataTypeTok{na.rm =} \NormalTok{T) }\CommentTok{# each col is a diff h, max over these}
      \NormalTok{index <-}\StringTok{ }\KeywordTok{which.max}\NormalTok{(V1[j,])  }\CommentTok{# store index so we can recover h's }
      \KeywordTok{c}\NormalTok{(value, index) }\CommentTok{# returns both profit value & index of optimal h.  }
    \NormalTok{\})}
    \CommentTok{# Sets V[t+1] = max_h V[t] at each possible state value, x}
    \NormalTok{V <-}\StringTok{ }\NormalTok{out[}\DecValTok{1}\NormalTok{,]                        }\CommentTok{# The new value-to-go}
    \NormalTok{D[,OptTime-time}\DecValTok{+1}\NormalTok{] <-}\StringTok{ }\NormalTok{out[}\DecValTok{2}\NormalTok{,]       }\CommentTok{# The index positions}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Replicate stochastic simulations
\item
  Sensitivity analysis (Figure 4).
\end{itemize}

\subsubsection{Example in a bistable system}

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape
parameter $\sigma_g$. We start with an example in which the parameters
are $r =$ 2, $K =$ 10, $C =$ 5, and $\sigma_g =$ 0.05.

Both parametric and nonparametric approaches will require some training
data on which to base their model of the process. We generate the
training data under the model described in Eq 1 for 35 time steps, under
a known but not necessarily optimal sequence of harvest intensities,
$h_t$. For simplicity we imagine a fishery that started from zero
harvest pressure and has been gradually increasing the harvest.
(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

Which estimates a Ricker model with $r =$

\begin{verbatim}

Error in eval(expr, envir, enclos) : object 'alt' not found
\end{verbatim}

, $K =$

\begin{verbatim}

Error in eval(expr, envir, enclos) : object 'alt' not found
\end{verbatim}

, and the Allen Allele model with $r =$

\begin{verbatim}

Error in eval(expr, envir, enclos) : object 'est' not found
\end{verbatim}

, $K =$

\begin{verbatim}

Error in eval(expr, envir, enclos) : object 'est' not found
\end{verbatim}

and $C =$

\begin{verbatim}

Error in eval(expr, envir, enclos) : object 'est' not found
\end{verbatim}

.

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Multiple species * Online learning * Multiple step-ahead predictions *
  Explicitly accomidating additional uncertainties * Improving inference
  of optimal policy from the GP
\end{itemize}

\texttt{\{r echo=FALSE, results="asis"\} \#bibliography("html")}

@@

Allen, Linda J. S., Jesse F. Fagan, Göran Högnäs, and Henrik Fagerholm.
2005. ``Population extinction in discrete-time stochastic population
models with an Allee effect.'' \emph{Journal of Difference Equations and
Applications} 11 (4-5) (apr): 273--293.
doi:10.1080/10236190412331335373.

@@

Cressie, Noel, Catherine a Calder, James S. Clark, Jay M. Ver Hoef, and
Christopher K. Wikle. 2009. ``Accounting for uncertainty in ecological
analysis: the strengths and limitations of hierarchical statistical
modeling.'' \emph{Ecological Applications} 19 (3) (apr): 553--70.

Geritz, Stefan a. H., and Eva Kisdi. 2011. ``Mathematical ecology: why
mechanistic models?'' \emph{Journal of mathematical biology} (dec).
doi:10.1007/s00285-011-0496-3.

Levins, Richard. 1966. ``The strategy of model building in population
biology.'' \emph{American Scientist} 54 (4): 421--431.

@@

Munch, Stephan B., Melissa L. Snover, George M. Watters, and Marc
Mangel. 2005. ``A unified treatment of top-down and bottom-up control of
reproduction in populations.'' \emph{Ecology Letters} 8 (7) (may):
691--695. doi:10.1111/j.1461-0248.2005.00766.x.

@@

Reed, William J. 1979. ``Optimal escapement levels in stochastic and
deterministic harvesting models.'' \emph{Journal of Environmental
Economics and Management} 6 (4) (dec): 350--363.
doi:10.1016/0095-0696(79)90014-7.

Scheffer, Marten, Stephen R. Carpenter, J. A. Foley, C. Folke, and B.
Walker. 2001. ``Catastrophic shifts in ecosystems.'' \emph{Nature} 413
(6856) (oct): 591--6. doi:10.1038/35098000.
